AA: The Algorithmic Autoregulation (Distributed Software Development) Methodology

### 2013-06-05 paper.tex
final version for wsl2013

To address these matters is the purpose of
AA methodology and an associated software system
for coordinating distributed team work, tackling the disadvantages
of GSD. Team members take on an egalitarian role, and stay
voluntarily \textit{logged} in the system for part of their time
(e.g.\ 2 hours per day), during which they log a periodical short text
sentence or \emph{microlog} --- similar to a `tweet' from Twitter --- as the
status of their activity. Logging is carried out using a series of
easy UI alternatives: unix shell
commands, native GUI or Web page, conventional social network posts, or chat messages to a log bot
listening to IRC, GTalk, G+, and others.  These ``microblog sentences'' are
publicly aggregated and validated by other team members. Through AA, the
community has a
methodology and an associated system to help
implement and validate the activities of a distributed software team. It implicitly
legitimizes financial support for the expansion of the activity of a
distributed development team. The AA methodology is specially useful for
coordinating distributed and decentralized team work, providing effective means
to asynchronously update different team members without the need for synchronous
unproductive meetings.

@@

There has been a large amount of research in methodologies to deal with
distributed teams of developers. Although this paper focuses on GSD, some of
its principles could also be adapted for smaller teams of developers working at
the same place, timezone and with minor cultural differences, depending on the
specific context and demands.  Moreover, `distributed development' is generally
tought of as being global which is not true. For instance, AA has been
effectively applied to a team whose members live in the same city but work at
different timeframes and locations, see Section~\ref{results}.  Even smaller
groups of developers working on the same building could use GSD methodologies
(or an adapted subset) to their benefit, \eg, to account for different work
habits, minimize formal meetings, document work process and history, and so on.
A thorough survey of these methodologies is beyond the scope of this paper; this
section presents but a brief overview.

@@

Second, cultural distance encompasses differences on organizational and
natural culture. Spoken language, unit and ethnic values are common
forms of such distance. Some companies prefer to allocate development
units in foreign locations with minimal cultural distance (\eg, an
American company may prefer Ireland due to spoken language
similarity~\cite{carmel2001}). Third, the temporal distance that
hampers synchronous communications like telephone or
videoconferences. Units of developers working on different time-zones
are concerned with managing of their agenda guided by this temporal
distance.

@@

Developers optionally record a brief \emph{video screencast} at the end of the
session summarizing what has been done, explaining his goals and challenges
in his own words and showing his most important results. This is very similar to
the video logging system in the movie Avatar, although it is clear, from
July 2011 git repositories and online wiki, that we have used this powerful concept in AA independently of any major mass midia incident. Moreover, screencasting differs
from general videologging in that it typically captures actual workflow on the
computer screen. Screencasting, combined with the textual log of the AA Session,
renders the final report more understandable for the individual developer
himself and to other people searching for information about his production.
% TODO: talvez figura com diagraminha explicando uma semana tipica,
% e cada sessao AA com um bloco dividido nos 15 min, e no inicio do bloco tb
% podemos listar no diagrama ``developer reads felow AA logs and lists his
% assigned tickets before he starts''

@@

All AA reports made by the developers are ultimatelly sent to a Web server and
become publically aggregated on a Website called pAAnel. It is then possible for a manager or another
developer to follow very closely the work of a given developer, nearly real-time,
reading each of his small reports or micrologs of what he is working on.

Another possibility is to check older sessions to check when certain tasks
were carried out and the comments of the developer about the \emph{process}.
Since each AA microlog happens in a very short timeframe of work, the information
about what was was done -- specially \emph{how} it was done -- becomes very easy to
understand, as opposed to having a long report in the end of a session.

@@

No set bosses or leaders are required in an ideal aplication the AA methodology;
in practice, the role of bosses basically disappears or is greatly alleviated --
hence the name Algorithmic \emph{Autoregulation} and other implicit
interpretations to the AA acronym and symbol. The primary mechanism to achieve
AA is peer coordination and social behavior, be it deliberate or implicit. In
order to prevent spamming and to improve the overall quality of AA reports, each
AA session must be validated by another developer. More specifically, all
reports are read by someone that will mark them collectively as `valid' or
`invalid' and may optionally write commentaries about the specific session and
quality of micrologs. The developer in charge of validating any given session is
randomly assigned by the AA Web server, which sends an email to the chosen
developer with an URL to a validation interface.

@@

Easy and effective GSD team management is the main purpose of the AA
methodology. The proposed methodology was applyed to a group of 9 developers in
July of 2011, 3 of which are coauthors of the present work, on what was named Lab
Macambira.\footnote{LabMacambira.sf.net: \url{http://labmacambira.sf.net}.} The
main objective of the team was to work on an array of strategic free software
projects in the broad audiovisual and web categories, contributing directly to
their development, submitting bug patches or committing new features to their
source code.

@@

\begin{table}
  \caption{Free and open software projects that received contributions from
  successfully initiated developers  or `Macambiras' that worked under the AA
  methodology. The first column lists applications to which contributions were
  officially accepted and whose development process was tracked using AA. The
  second column shows the pseudonym of the committers. At Lab Macambira it is
  common practice to use pseudonyms in AA as identification in order to enhance
  privacy.}
  \begin{center}
  \small\begin{tabular}{|l|l|}
        \hline
        Application           & ``Commiters''                       \\
        \hline \hline
        Mozilla Firefox       & daneoshiga, bzum                    \\
        Evince                & hick209, bzum, marcicano, mquasar   \\
        BePDF / Xpdf          & marcicano                           \\
        Ekiga                 & flecha                              \\
        Empathy               & fefo                                \\
        Lib Folks (Telepathy) & kamiarc                             \\
        Scilab                & v1z, humannoise                     \\
        VxL                   & v1z                                 \\
        ImageMagick           & v1z                                 \\
        OpenOffice            & hick209                             \\
        Puredata              & v1z, automata, greenkobold, gilson, bzum \\
        Puredata OpenCV       & v1z                                 \\
        Puredata GEM          & v1z, fefo, hick209                  \\
        Puredata PDP          & v1z, fefo, hick209                  \\
        ChucK                 & rfabbri, automata                   \\
        ChucK MiniAudicle     & rfabbri, automata                   \\
        WebRTC                & automata                            \\
        OSC-Web               & automata                            \\
        Web-PD-GUI            & automata                            \\
        Live-Processing       & automata                            \\
        ChucK-Wiimote         & automata                            \\
        Audiolet              & automata                            \\
        Extempore             & automata                            \\
        \hline
    \end{tabular}
    \end{center}
    \label{tabela:contribuicoes}
\end{table}

@@

While using the AA system, developers learned to work asynchronously with others
and got used to the habit of periodically updating their status on their
projects. Each programmer was given the chance to work with considerable
freedom, in any place and time of preference. The strictest required
responsibility was that of using AA for at least one sesssion of 2h per day,
while working on the agreed upon tasks.  The online pAAnel allowed each
developer to quickly grasp activities from others while avoiding interrupting them, a
process further aided by the screencasts. Adjustments to the task deadlines and
milestones (which were managed in Trac) were done based on verified progress of
individuals and labMacambira.sf.net team as a whole. The various newcommers
benefitted from a fast learning team by the use of AA as a flexible and simple
transparency system. Updates from the team were not transmitted on a
person-to-person basis, but rather on a person-to-team basis based on the
available online progress information.

\begin{table}
    \caption{Software projects created by Lab Macambira since July
      2011 with a short description and the technologies involved (\eg,
      programming languages or frameworks). It is
      interesting to note the heterogeneity of projects and their areas
      of application.}
    \footnotesize{\begin{tabular}{|l|p{5cm}|l|}
        \hline
        Application & Description & Technologies \\ 
        \hline \hline
        AA            & Algorithmic Autoregulation      & Python, PHP \\
        \hline
        \'{A}gora Communs & System for on-line deliberations & PHP \\
        \hline
        SIP           & Scilab Image Processing toolbox & C, Scilab \\
        \hline
        Animal        & An Imaging Library              & C \\
        \hline
        TeDi          & Test Framework for Distance Transform
        Algorithms & C, Shellscript, Scilab \\
        \hline
        Macambot      & Multi-use IRC Bot               & Python \\
        \hline
        Confer\^{e}ncia Permanente & Platform for the permanent
        conference of the rights of minors & PHP, JavaScript \\
        \hline
        CPC           & Center for accounting of the Brazilian culture
        representation groups & Python, Django \\
        \hline
        Timeline      & Interactive time lines on the Web & JavaScript
        \\
        \hline
        Imagemap      & Interactive marking for on-line photos &
        JavaScript \\
        \hline
        ABT           & Program for real-time execution and musical
        rhythmic analysis & Python \\
        \hline
        EKP           & Emotional Kernel Panic & Python, ChucK \\
        \hline
        SOS           & Aggregation and diffusion of popular and native
        knowledge about health & Python, Django \\
        \hline
        Creative Economy & Platform for creative, collaborative and
        solidarity economy of the culture hubs and cultural entities &
        Python, Django \\
        \hline
        OpenID Integration & Adaptations to existing software for
        unified login through OpenID & PHP \\
        \hline
        pAAinel & Dashboard for the real-time visualization of Lab
        Macambira activity & Python, Django \\
        \hline
        Georef & Collection of scripts to be used as reference, which
        aims to be a GIS platform to map public data of use to
        citizens & Python, Django \\
        \hline
        AirHackTable & Software for an instrument which generates
        sound from flying origami tracked by webcams & Puredata,
        C++, Scilab \\
        \hline
        \end{tabular}}
    \label{tabela:criados}
\end{table}

@@

\section*{Acknowledgments}
Authors acknowledge the financial support from Pontao Nos Digitais, and
Ricardo Fabbri acknowledges support from FAPERJ/Brazil 111.852/2012.
Authors also thank AA: the present research and even this
manuscript was written using AA. The complete log is on-line at
\url{www.pulapirata.com/skills/aa}. Finally, authors are also grateful to IFSC/USP, IPRJ/UERJ, IFT/UNESP and all AA users and collaborators, especially those who coded AA hacks for use through shell and bot, and those who coded different Web interfaces in use.

>>>>

To address these matters is the purpose of
AA methodology and an associated software system
for coordinating distributed team work, tackling the disadvantages
of GSD. Team members take on an egalitarian role, and stay
voluntarily \textit{logged} in the system for part of their time
(e.g.\ 2 hours per day), during which they log a periodical short text
sentence or \emph{microlog} --- similar to a `tweet' from Twitter --- as the
status of their activity. Logging is carried out using a series of
easy UI alternatives: UNIX shell
commands, native GUI or Web page, conventional social network posts, or chat messages to a log bot
listening to IRC, GTalk, G+, and others.  These ``microblog sentences'' are
publicly aggregated and validated by other team members. Through AA, the
community has a
methodology and an associated system to help
implement and validate the activities of a distributed software team. It implicitly
legitimizes financial support for the expansion of the activity of a
distributed development team. The AA methodology is specially useful for
coordinating distributed and decentralized team work, providing effective means
to asynchronously update different team members without the need for synchronous
unproductive meetings.

@@

There has been a large amount of research in methodologies to deal with
distributed teams of developers. Although this paper focuses on GSD, some of
its principles could also be adapted for smaller teams of developers working at
the same place, timezone and with minor cultural differences, depending on the
specific context and demands.  Moreover, `distributed development' is generally
thought of as being global which is not true. For instance, AA has been
effectively applied to a team whose members live in the same city but work at
different timeframes and locations, see Section~\ref{results}.  Even smaller
groups of developers working on the same building could use GSD methodologies
(or an adapted subset) to their benefit, \eg, to account for different work
habits, minimize formal meetings, document work process and history, and so on.
A thorough survey of these methodologies is beyond the scope of this paper; this
section presents a brief overview.

@@

Second, cultural distance encompasses differences on organizational and
natural culture. Spoken language, unit and ethnic values are common
forms of such distance. Some companies prefer to allocate development
units in foreign locations with minimal cultural distance (\eg, an
American company may prefer Ireland due to spoken language
similarity~\cite{carmel2001}). Third, the temporal distance that
hampers synchronous communications like telephone or
video conferences. Units of developers working on different time-zones
are concerned with managing of their agenda guided by this temporal
distance.

@@

Developers optionally record a brief \emph{video screencast} at the end of the
session summarizing what has been done, explaining his goals and challenges
in his own words and showing his most important results. This is very similar to
the video logging system in the movie Avatar, although it is clear, from
July 2011 git repositories and online wiki, that we have used this powerful concept in AA independently of any major mass media incident. Moreover, screencasting differs
from general videologging in that it typically captures actual workflow on the
computer screen. Screencasting, combined with the textual log of the AA Session,
renders the final report more understandable for the individual developer
himself and to other people searching for information about his production.
% TODO: talvez figura com diagraminha explicando uma semana tipica,
% e cada sessao AA com um bloco dividido nos 15 min, e no inicio do bloco tb
% podemos listar no diagrama ``developer reads felow AA logs and lists his
% assigned tickets before he starts''

@@

All AA reports made by the developers are ultimately sent to a Web server and
become publicly aggregated on a Website called pAAnel. It is then possible for a manager or another
developer to follow very closely the work of a given developer, nearly real-time,
reading each of his small reports or micrologs of what he is working on.

Another possibility is to check older sessions to check when certain tasks
were carried out and the comments of the developer about the \emph{process}.
Since each AA microlog happens in a very short timeframe of work, the information
about what was done -- specially \emph{how} it was done -- becomes very easy to
understand, as opposed to having a long report in the end of a session.

@@

No set bosses or leaders are required in an ideal application the AA methodology;
in practice, the role of bosses basically disappears or is greatly alleviated --
hence the name Algorithmic \emph{Autoregulation} and other implicit
interpretations to the AA acronym and symbol. The primary mechanism to achieve
AA is peer coordination and social behavior, be it deliberate or implicit. In
order to prevent spamming and to improve the overall quality of AA reports, each
AA session must be validated by another developer. More specifically, all
reports are read by someone that will mark them collectively as `valid' or
`invalid' and may optionally write commentaries about the specific session and
quality of micrologs. The developer in charge of validating any given session is
randomly assigned by the AA Web server, which sends an email to the chosen
developer with an URL to a validation interface.

@@

Easy and effective GSD team management is the main purpose of the AA
methodology. The proposed methodology was applied to a group of 9 developers in
July of 2011, 3 of which are coauthors of the present work, on what was named Lab
Macambira.\footnote{LabMacambira.sf.net: \url{http://labmacambira.sf.net}.} The
main objective of the team was to work on an array of strategic free software
projects in the broad audiovisual and web categories, contributing directly to
their development, submitting bug patches or committing new features to their
source code.

@@

\begin{table}
  \caption{Free and open software projects that received contributions from
  successfully initiated developers  or `Macambiras' that worked under the AA
  methodology. The first column lists applications to which contributions were
  officially accepted and whose development process was tracked using AA. The
  second column shows the pseudonym of the ``committers''. At Lab Macambira it is
  common practice to use pseudonyms in AA as identification in order to enhance
  privacy.}
  \begin{center}
  \small\begin{tabular}{|l|l|}
        \hline
        Application           & ``Committers''                       \\
        \hline \hline
        Mozilla Firefox       & daneoshiga, bzum                    \\
        Evince                & hick209, bzum, marcicano, mquasar   \\
        BePDF / Xpdf          & marcicano                           \\
        Ekiga                 & flecha                              \\
        Empathy               & fefo                                \\
        Lib Folks (Telepathy) & kamiarc                             \\
        Scilab                & v1z, humannoise                     \\
        VxL                   & v1z                                 \\
        ImageMagick           & v1z                                 \\
        OpenOffice            & hick209                             \\
        Puredata              & v1z, automata, greenkobold, gilson, bzum \\
        Puredata OpenCV       & v1z                                 \\
        Puredata GEM          & v1z, fefo, hick209                  \\
        Puredata PDP          & v1z, fefo, hick209                  \\
        ChucK                 & rfabbri, automata                   \\
        ChucK MiniAudicle     & rfabbri, automata                   \\
        WebRTC                & automata                            \\
        OSC-Web               & automata                            \\
        Web-PD-GUI            & automata                            \\
        Live-Processing       & automata                            \\
        ChucK-Wiimote         & automata                            \\
        Audiolet              & automata                            \\
        Extempore             & automata                            \\
        \hline
    \end{tabular}
    \end{center}
    \label{tabela:contribuicoes}
\end{table}

@@

While using the AA system, developers learned to work asynchronously with others
and got used to the habit of periodically updating their status on their
projects. Each programmer was given the chance to work with considerable
freedom, in any place and time of preference. The strictest required
responsibility was that of using AA for at least one session of 2h per day,
while working on the agreed upon tasks.  The online pAAnel allowed each
developer to quickly grasp activities from others while avoiding interrupting them, a
process further aided by the screencasts. Adjustments to the task deadlines and
milestones (which were managed in Trac) were done based on verified progress of
individuals and labMacambira.sf.net team as a whole. The various newcomers
benefited from a fast learning team by the use of AA as a flexible and simple
transparency system. Updates from the team were not transmitted on a
person-to-person basis, but rather on a person-to-team basis based on the
available online progress information.

\begin{table}
    \caption{Software projects created by Lab Macambira since July
      2011 with a short description and the technologies involved (\eg,
      programming languages or frameworks). It is
      interesting to note the heterogeneity of projects and their areas
      of application.}
    \footnotesize{\begin{tabular}{|l|p{5cm}|l|}
        \hline
        Application & Description & Technologies \\ 
        \hline \hline
        AA            & Algorithmic Autoregulation      & Python, PHP \\
        \hline
        \'{A}gora Communs & System for on-line deliberations & PHP \\
        \hline
        SIP           & Scilab Image Processing toolbox & C, Scilab \\
        \hline
        Animal        & An Imaging Library              & C \\
        \hline
        TeDi          & Test Framework for Distance Transform
        Algorithms & C, Shellscript, Scilab \\
        \hline
        Macambot      & Multi-use IRC Bot               & Python \\
        \hline
        Confer\^{e}ncia Permanente & Platform for the permanent
        conference of the rights of minors & PHP, JavaScript \\
        \hline
        CPC           & Center for accounting of the Brazilian culture
        representation groups & Python, Django \\
        \hline
        Timeline      & Interactive time lines on the Web & JavaScript
        \\
        \hline
        Imagemap      & Interactive marking for on-line photos &
        JavaScript \\
        \hline
        ABT           & Program for real-time execution and musical
        rhythmic analysis & Python \\
        \hline
        EKP           & Emotional Kernel Panic & Python, ChucK \\
        \hline
        SOS           & Aggregation and diffusion of popular and native
        knowledge about health & Python, Django \\
        \hline
        Creative Economy & Platform for creative, collaborative and
        solidarity economy of the culture hubs and cultural entities &
        Python, Django \\
        \hline
        OpenID Integration & Adaptations to existing software for
        unified login through OpenID & PHP \\
        \hline
        pAAnel & Dashboard for the real-time visualization of Lab
        Macambira activity & Python, Django \\
        \hline
        Georef & Collection of scripts to be used as reference, which
        aims to be a GIS platform to map public data of use to
        citizens & Python, Django \\
        \hline
        AirHackTable & Software for an instrument which generates
        sound from flying origami tracked by webcams & Puredata,
        C++, Scilab \\
        \hline
        \end{tabular}}
    \label{tabela:criados}
\end{table}

@@

\section*{Acknowledgments}
Authors acknowledge the financial support from Pont\~{a}o N\'{o}s Digitais, and
Ricardo Fabbri acknowledges support from FAPERJ/Brazil 111.852/2012.
Authors also thank AA: the present research and even this
manuscript was written using AA. The complete log is on-line at
\url{www.pulapirata.com/skills/aa}. Finally, authors are also grateful to IFSC/USP, IPRJ/UERJ, IFT/UNESP and all AA users and collaborators, especially those who coded AA hacks for use through shell and bot, and those who coded different Web interfaces in use.
